## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packages
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv",stringsAsFactors = T)

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") 
class(USGS.flow.data$datetime)

```


## Visualization for Data Exploration

Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar()

ggplot(na.omit(USGS.flow.data), aes(x = discharge.mean.approval)) +
  geom_bar()

```

### Histogram (function: geom_histogram)

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

#
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500))
  
#
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))

```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "darkgray") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) +
  scale_x_continuous(limits = c(0, 10))

#
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin)

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: 
Each of these types of plots are best suited for working with difference classes of variables. For example, bar plots work well for counts of categorical variables while histograms show distributions of numerical data. It also depends on what type of question you're trying to explore. The box-and-whisker or violin plots work well for visualizing distributions while the scatter plots work well for visualizing trends or relationships. 

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: 
I don't have experience working with stream data, so I learned what discharge and gage height mean in terms of working with water/stream data. I also learned about how to work with historical data going past where R tends to automatically differentiate between the 20th and 21st centuries. Generally the different graph types show how the discharge mean or the gage height are distributed and the relationship between the two. You can see where the typical values of the variable are and where the outliers stand out. 


How can multiple options for data exploration inform our understanding of our data?

> Answer: 
You can ask a lot of different questions while exploring data - from visualizing trends, relationships, or distributions to making sure the variables are in the appropriate class for analysis and column names makes sense to human readers. 


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: 
The pattern that stands out to me the most is the relationship between discharge and gage height, as visualized in the scatter plot and the group cut width box plot. As discharge increases, the gage height increases. This makes sense becuase if discharge is increasind and more water is flowing through the stream, you would expect the height of the water surface above the gage to increase due to the increased water flow/volume. 


