---
title: "Assignment 5: Data Visualization"
author: "Elizabeth Good"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
# load in needed packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

# verify home directory 
here()

# read in processed data 
PeterPaul.chem.nutrients <- 
  read.csv(
    here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
           stringsAsFactors = TRUE)

NEON.NIWO.litter <- 
  read.csv(
    here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
           stringsAsFactors = TRUE)

#2 
class(PeterPaul.chem.nutrients$sampledate)
class(NEON.NIWO.litter$collectDate)
# they are currently all of class factor

PeterPaul.chem.nutrients$sampledate <- 
  ymd(PeterPaul.chem.nutrients$sampledate)
NEON.NIWO.litter$collectDate <- 
  ymd(NEON.NIWO.litter$collectDate)

class(PeterPaul.chem.nutrients$sampledate)
class(NEON.NIWO.litter$collectDate)
# now they are all of class "Date"
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
new_theme <- theme_bw() +
  theme(axis.text = element_text(color = "navy", 
                                 size = 12), 
        axis.title = element_text(color = "gray40",
                                  size = 12),
        plot.title = element_text(color = "gray40", 
                                  face = "bold",
                                  hjust = 0.5),
        legend.position = "bottom")

theme_set(new_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
phosphorus_phosphate_plot <- PeterPaul.chem.nutrients %>% 
  ggplot(mapping = aes(x = po4,
                       y = tp_ug,
                       color = lakename)) + 
  ggtitle("Phosphorus and Phosphate at Peter and Paul Lakes") +
  scale_y_continuous(name = "Total Phosphorus (ug)",
                     limits = (c(-5,160))) +
  scale_x_continuous(name = "Phosphate (ug)",
                     limits = (c(-5,60))) +
  labs(color = 'Lake') +
  geom_point(alpha = 0.5) +
  scale_color_brewer(palette = "Set1") +
  geom_smooth(method = lm, color = "black")

print(phosphorus_phosphate_plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
a <- 
  ggplot(PeterPaul.chem.nutrients, 
            aes(x = factor(month, 
                           levels = c(1:12),
                           labels = c(month.abb)), 
                y = temperature_C)) +
  ggtitle("Temp and Nutrient Trends at Peter and Paul Lakes") +
  geom_boxplot(aes(color = lakename)) +
  labs(color = 'Lake') +
  scale_y_continuous(name = "Temp (deg C)") +
  scale_x_discrete(name = "Month")

# print(a)

b <-
  ggplot(PeterPaul.chem.nutrients, 
         aes(x = factor(month, 
                        levels = c(1:12),
                        labels = c(month.abb)), 
             y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(color = 'Lake') +
  scale_y_continuous(name = "Total P (ug)") +
  scale_x_discrete(name = "Month")

# print(b)

c <-
  ggplot(PeterPaul.chem.nutrients, 
         aes(x = factor(month, 
                        levels = c(1:12),
                        labels = c(month.abb)), 
             y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(color = 'Lake') +
  scale_y_continuous(name = "Total N (ug)") +
  scale_x_discrete(name = "Month")

# print(c)

combined_plots <- 
  plot_grid(a + theme(legend.position = "none"), 
            b + theme(legend.position = "none"), 
            c + theme(legend.position = "none"), 
            nrow = 3, 
            align = 'v')

legend <- 
  get_legend(a)

combined_plots_legend <- 
  plot_grid(combined_plots, 
            legend, 
            ncol = 1, 
            rel_heights = c(1, .1))

print(combined_plots_legend)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 
There is a noticeable rise in temperature during summer months across both lakes. There doesn't appear to be a lot of seasonal variability in total phosphorus or nitrogren, but it does look like Peter Lake has slightly higher concentrations of both nutrients compared to Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
needle_plot_color <- 
  ggplot(NEON.NIWO.litter %>%
           filter(functionalGroup == "Needles"), 
         aes(x = collectDate, 
             y = dryMass, 
             color = nlcdClass)) +
  ggtitle("Needle Dry Mass") +
  geom_point(alpha = 1, 
             size = 2.5) +
  scale_x_date(name = "Date", 
               date_breaks = "3 months", 
               date_labels = "%b %Y") +
  scale_y_continuous(name = "Dry Mass (g)") +
  labs(color = "NLCD Class")+
  scale_color_brewer(palette = "Paired") +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))

print(needle_plot_color)

#7
needle_plot_facet <- 
  ggplot(NEON.NIWO.litter %>%
           filter(functionalGroup == "Needles"), 
         aes(x = collectDate, 
             y = dryMass)) +
  ggtitle("Needle Dry Mass") +
  facet_wrap(vars(nlcdClass), 
             nrow = 3) +
  geom_point(alpha = 0.8,
             size = 2,
             color = "skyblue3") +
  scale_x_date(name = "Date", 
               date_breaks = "3 months", 
               date_labels = "%b %Y") +
  scale_y_continuous(name = "Dry Mass (g)") +
  labs(color = "NLCD Class")+
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust=1))

print(needle_plot_facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
Personally, I think the second plot separated into facets instead of separated by color is a more effective visual. There's a lot of point overlap, making it difficult to distinguish classes in the first plot. Because the y axes stay the same across the faceted plots, it's easy to see at a glance which classes are similar and which have a greater distribution without the same difficulty trying the distinguish point colors. For example, it's easy to see in the second plot that in 2016 and 2017 there's a greater distribution in the evergreen and grassland classes, while the point distribution looks very similar across all three classes in 2018 and 2019. I think this is harder to see at a gland in the first plot when you're trying to see which colors are present in a big cluster of points. 