---
title: "Assignment 2: Coding Basics"
author: "Elizabeth Good"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
thirty_sequence <- seq(1, 30, 3)
# I'm creating a sequence of numbers from 1 to 30 increasing by 3 and giving this sequence 
# the name "thirty_sequence".
#2. 
mean(thirty_sequence) # this computes the mean of thirty_sequence, which equals 14.5
median(thirty_sequence) # this computes the median of thirty_sequence which equals 14.5
#3. 
mean(thirty_sequence)>median(thirty_sequence) # this is asking if the mean of 
# thirty_sequence is greater than the median of thirty_sequence, to which the answer is FALSE 
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5. #6. 
names <- c("Sue", "Rob", "Chris", "Greg") # type of vector is character
scores <- c( 87, 90, 99, 83) # type of vector is numeric
pass <- c(TRUE, TRUE, TRUE, TRUE) # type of vector is logical 
#7. 
student_stats <- data.frame(names, scores, pass)
#8. 
library(dplyr)
student_stats <- student_stats %>%
  rename(student_name = "names",
         test_score = "scores",
         pass_true_or_false = "pass")
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
A matrix can only consist of a single class of data while a data frame can contain a variety of classes of data. In this case, the data frame has numerical, logical, and character types of data which would not be able to coexist in a matrix. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#10.
pass_or_fail <- function(x){
  y <- ifelse(x < 50, "FALSE", "TRUE")
  print(y)
}

# pass_or_fail2 <- function(x) {
#   if(x > 50) {
#     print(TRUE)
#   } 
#   else {
#     print(FALSE)
#   }
# }
#11. 
pass_or_fail(student_stats$test_score)
# pass_or_fail2(student_stats$test_score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
Using `ifelse` works because it can accept a vector. `If...else` will not work with a vector input. It gives an error that the condition has a length greater than 1. The if function can only take individual elements and not a vector. Note that in the code above I tried both methods and commented out the function and test that didn't work. 